Executive Constraints and Civil Conflict Onset

### 2019-11-02 paper.Rmd
WIP: theory section

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
documentclass: scrartcl
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
    html_document:
        css: assets/sakura.css
        smart: true
        toc: true
---

```{r include = F, results = "hide"}
# Start by setting up our R env
knitr::opts_chunk$set(echo = F, eval = F)

library(dplyr)
library(precrec)
library(scales)
library(thesis.utils)

@@

\newpage

# Appendix

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
documentclass: scrreprt
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
    html_document:
        css: assets/sakura.css
        smart: true
        toc: true
---

[//]: # Setup R environment for entire doc
```{r include = F, results = "hide"}
knitr::opts_chunk$set(echo = F, eval = F)

library(dplyr)
library(thesis.utils)
```

[//]: # Force TOC on separate page
\newpage

# Introduction {-}

# Theoretical Overview {-}

## Institutions and Civil Conflict

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" (cite Hodgson). More informally, we can consider
them to be the "rules of the game", encompassing for example codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents.^[None of which is meant to
imply that institutions stand external to actor perceptions and
preferences as unchanging features. Rather, institutions are social
constructs subject to changing normative values and able to be
reinforced or undermined by behavioural feedback loops. This latter
point is especially interesting within the context of civil conflict
considering how the development of various institutions and conflict
incidence can be jointly determined; however, it remains outside the
scope of this paper.] Furthermore, over time, the habits formed from
these systems create expectations of stable patterns of behaviour
thereby shaping the interactions between actors by reducing
uncertainty (cite North). Although there is a wide variety of
institutions, the present focus will be on the type of institutions
which govern political behaviour.

This exposition is relevant to the study of civil conflict for both
motivation-based and opportunity-based explanatory accounts. To the
extent that the onset of civil conflict is a longer process featuring
waves of escalation and de-escalation rather than a single one-off
event (Florea), institutions are a key influence on the constitutive
interactions between state and non-state actors. For example, they can
shift the opportunity costs of violence altering the mobilization
capacity of would be rebel groups. Institutions can also ensure
credible commitments towards negotiated settlements, assuming that
actors are rational and able to internalize the cost of violence such
that a bargained solution is preferable to fighting (Blattman &
Fearon). And, their quality can shape the capacity of the state to
suppress or co-opt potential opposition. None of these examples are
meant to form an exhaustive list. Instead the point is to demonstrate
the number of different ways in which institutions can enter into the
theoretical discussions about civil conflict.

Unsurprisingly, insofar as the role of institutions has been examined
by previous literature the focus has largely been on the broader
notion of democracy, which at a minimum is a system regulating access
to political power through popular representation determined by free
and competitive elections (Schumpter). Democracy has been hypothesised
to have a pacifying domestic effect according to two predominant
mechanisms. The first is centered on the concept of relative
deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence (Gurr 1970). [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means (Hegre).

[^gurr]: Gurr explains value expectations as being that which we think
    we are entitled to and value capabilities as that which we think
    we are capable of attaining. As Gurr then states, relative
    deprivation is the perceived discrepancy between the two. The link
    to collective violence, however, is conditional on the discontent
    being politicized, by which it is meant that the source of the
    discrepancy is identified and a normative and utilitarian
    justification is given for violent action, before finally being
    actualized.  Quantitative studies tend to focus on the first
    stage, the propensity for political regimes to generate
    discontent, given the difficultly of measuring the variables that
    may affect politicisation and actualisation.

## Executive Constraints

# Empirics {-}

@@

# Conclusion {-}

# Appendix {-}
